Skew-symmetric entropy stable discontinuous Galerkin formulations with application to hybrid meshes

### 2018-06-20 docs/skewform/main.tex
removing old okl, adding okl kernel building

\documentclass[preprint,10pt]{elsarticle}
%\documentclass[preprint,10pt]{article}
%\documentclass[review]{siamart0216}
%\documentclass{siamart0216}

\usepackage{fullpage}
\usepackage[colorlinks=true]{hyperref}

\usepackage{amsmath,amssymb,amsfonts,amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{lemma}
\newtheorem{lemma}{Lemma}
\theoremstyle{corollary}
\newtheorem{corollary}{Corollary}
\newtheorem*{remark}{Remark}
\theoremstyle{theorem}
\newtheorem{theorem}{Theorem}
\theoremstyle{assumption}
\newtheorem{assumption}{Assumption}

@@

%\date{}
%\author{Jesse Chan}
%\title{On discretely entropy conservative discontinuous Galerkin methods}
\graphicspath{{./figs/}}

@@

%\maketitle

\begin{frontmatter}
\title{Skew-symmetric formulations of entropy stable discontinuous Galerkin methods with applications to hybrid and non-conforming meshes. }

\author[rice]{Jesse Chan\corref{cor1}}
\ead{Jesse.Chan@caam.rice.edu}
\address[rice]{Department of Computational and Applied Mathematics, Rice University, 6100 Main St, Houston, TX, 77005}

\begin{abstract}
Entropy stable high order methods for nonlinear conservation laws satisfy an inherent discrete entropy inequality.  
\end{abstract}
\end{frontmatter}

@@

@@

The schemes in \cite{chan2017discretely} begin on the reference element $\hat{D}$ by approximating the solution in a degree $N$ polynomial basis $\LRc{\phi_j({\bm{x}})}_{i=1}^{N_p}$.  It is assumed that volume and surface quadrature rules $({\bm{x}}_i, w_i)$, $\LRp{{\bm{x}}^f_i,w^f_i}$ are provided for $\hat{D}$ and the surface $\partial \hat{D}$ of $\hat{D}$, respectively.  We make the following assumption on the volume quadrature rule:
\begin{assumption}
The volume quadrature rule is exact for polynomials of degree $(2N-1)$.
\label{ass:quad}
\end{assumption}
In addition to being related to the positive definiteness of the mass matrix, Assumption~\ref{ass:quad} ensures that, for any $u,v \in V^N$, the following integral is exactly approximated using quadrature
\[
\int_{\hat{D}} \pd{u(\hat{\bm{x}})}{\hat{x}_i} v(\hat{\bm{x}})\diff{\hat{\bm{x}}} = \sum_{k=1}^{N_q} w_k \pd{u(\hat{\bm{x}}_k)}{\hat{x}_i} v(\hat{\bm{x}}_k).  
\]
In other words, evaluating the integral of the derivative of $u$ and $v$ using quadrature is exact because the integrand is a polynomial of degree $(2N-1)$.

\note{TODO: Standardize hat notation for operators/normals.}

@@

Let $\bm{n}$ denote the outward unit normal $\bm{n}$ on $\hat{D}$, and let $\bm{n}_i$ denote the vector containing values of the $i$th component at all surface quadrature points.  The ``decoupled'' summation by parts operator $\bm{D}^i_N$ is defined as the block matrix involving both volume and surface quadratures
\begin{align}
\bm{D}^i_N  &= \LRs{
\begin{array}{cc}
\bm{D}^i_q - \frac{1}{2}\bm{V}_q \bm{L}_f {\rm diag}({\bm{n}}_i) \bm{V}_f\bm{P}_q &  \frac{1}{2}\bm{V}_q\bm{L}_f{\rm diag}({\bm{n}}_i)\\
-\frac{1}{2}{\rm diag}({\bm{n}}_i)\bm{V}_f\bm{P}_q & \frac{1}{2}{\rm diag}({\bm{n}}_i)
\end{array}}.  
\label{eq:DN}
\end{align}
When combined with projection and lifting matrices, $\bm{D}^i_N$ produces a high order polynomial approximation of $f\pd{g}{x_i}$. 
Let $f, g$ be differentiable functions, and let $\bm{f}_i = f(\bm{x}_i)$, $\bm{g}_i = g(\bm{x}_i)$ denote values of $f,g$ at volume and surface quadrature points.  Then,
\begin{gather}
f\pd{g}{x_i} \approx \LRs{\begin{array}{cc}
\bm{P}_q & \bm{L}_f\end{array}} {\rm diag}\LRp{\bm{f}}\bm{D}^i_N \bm{g}.  
\label{eq:dnapprox}
\end{gather}
The approximation can also be interpreted as solving a variational problem.  Let $n_i$ be the $i$th component of the unit normal on $\hat{D}$.  Then (\ref{eq:dnapprox}) is equivalent to finding $u(\bm{x}) \approx f\pd{g}{x_i}$ such that 
\[
\int_{\hat{D}} u(\bm{x})v(\bm{x})  = \int_{\hat{D}}{g\pd{\Pi_Nf}{x_i}v} + \int_{\partial \hat{D}}{(f-\Pi_Nf)\frac{\LRp{gv + \Pi_N(gv)}}{2}}n_i, \qquad \forall v\in V^N\LRp{\hat{D}}.
\]
If the surface quadrature rule is exact for polynomials of degree $2N$, then the decoupled SBP operator $\bm{D}^i_N$ also satisfies a summation by parts property.  
\begin{gather}
\bm{Q}^i_N = \bm{W}_N \bm{D}^i_N, \qquad \bm{W}_N = \LRp{\begin{array}{cc}\bm{W}&\\ & \bm{W}_f\end{array}},\label{eq:dsbp}\\ \bm{Q}^i_N+\LRp{\bm{Q}^i_N}^T = \LRp{\begin{array}{cc}\bm{0}&\\ & \bm{W}_f {\rm diag}(\bm{n}_i)\end{array}} = \bm{B}_N\nonumber
\end{gather}
By noting that $\bm{W}\bm{V}_q \bm{L}_q = \bm{P}_q^T\bm{V}_f^T\bm{W}_f$, the entries of $\bm{Q}^i_N$ can be re-written as
\begin{align}
\bm{Q}^i_N  &= \LRs{
\begin{array}{cc}
\bm{Q}_i - \frac{1}{2}\LRp{\bm{V}_f \bm{P}_q}^T  \bm{W}_f{\rm diag}({\bm{n}}_i) \bm{V}_f\bm{P}_q &  \frac{1}{2}\LRp{\bm{W}_f{\rm diag}({\bm{n}}_i)\bm{V}_f\bm{P}_q}^T\\
-\frac{1}{2}\bm{W}_f{\rm diag}({\bm{n}}_i)\bm{V}_f\bm{P}_q & \frac{1}{2}\bm{W}_f{\rm diag}({\bm{n}}_i)
\end{array}}, \qquad \bm{Q}_i = \bm{W}\bm{V}_q\bm{D}^i\bm{P}_q,
\label{eq:QN}
\end{align}
from which the SBP property (\ref{eq:dsbp}) is more apparent.

@@

\subsection{Loss of the SBP property for hybrid and non-conforming meshes}
While the SBP property has been used to derive entropy stable schemes, the SBP property is difficult or impossible to enforce in certain discretization settings, such as hybrid and non-conforming meshes arises.  This difficulty is a result of the choices of surface quadrature which naturally arise for these settings.

\begin{figure}
\centering
\begingroup
\captionsetup[subfigure]{width=.43\textwidth}
\subfloat[Insufficiently accurate surface quadrature on the triangle element.]{\includegraphics[width=.425\textwidth]{figs/hybrid2D.png}\label{subfig:hybrid1}}
\endgroup
\hspace{2em}
\subfloat[Incompatible surface quadrature on the quadrilateral element.]{\includegraphics[width=.425\textwidth]{figs/hybrid2D_GQ.png}\label{subfig:hybrid2}}
\caption{Examples of interface couplings which do not result in an SBP property and are not entropy stable.  On the left, the surface quadrature is a $(N+1)$ point GLL rule, and results in a loss of the SBP property on the triangle.  On the right, the surface quadrature is a $(N+1)$ point Gauss-Legendre rule, and results in a loss of the SBP property on the quadrilateral.  }
\label{fig:hybrid}
\end{figure}

@@

\begin{figure}[!h]
\centering
\begingroup
\captionsetup[subfigure]{width=.425\textwidth}
\subfloat[Coupling of nodes for a non-conforming interface (entropy stable but tighter coupling).]{\raisebox{0em}{\includegraphics[width=.425\textwidth]{figs/noncon.png}}\label{subfig:noncon2}}
\hspace{2em}
\subfloat[Surface quadrature rule induced by non-conforming refinement (no SBP property).]{\raisebox{-.3em}{\includegraphics[width=.45\textwidth]{figs/nonconQuad.png}}\label{subfig:noncon1}}
\endgroup
%\subfloat[3D hexahedra-pyramid coupling]{\includegraphics[width=.31\textwidth]{figs/hybrid3D.png}}
\caption{Illustration of two different entropy stable inter-element couplings for $h$ non-conforming meshes.  Each dashed red line indicates a dyadic flux computation required between two nodes.  The coupling terms in \cite{friedrich2017entropy} requires computing dyadic fluxes between \emph{each} pair of nodes on adjacent interfaces, while the proposed approach requires dyadic flux computations only between immediately adjacent nodes.   }
\label{fig:noncon}
\end{figure}

@@

\begin{lemma}
Let $v\in V^N$, and let the surface trace $\LRl{v}_{\partial \hat{D}} \in V^M\LRp{\partial \hat{D}}$ for some $M \leq N$. If the surface quadrature is exact for polynomials of degree $N+M$,  then for an arbitrary vector $\bm{u}_q$, 
\[
\bm{v}_q^T\bm{Q}_i \bm{u}_q = \bm{v}_q^T\LRp{ \LRp{\bm{V}_f\bm{P}_q}^T \bm{W}_f \diag{{\bm{n}}_i}\bm{V}_f\bm{P}_q - \bm{Q}_i^T}\bm{u}_q
\]
where $\bm{v}$ denotes the vector of values of $v$ of at both volume and surface quadrature points. 
\label{lemma:vsbp}
\end{lemma}
\begin{proof}
Recall that $\bm{Q}_i = \bm{W} \bm{V}_q \bm{D}^i\bm{P}_q$.  Since the volume quadrature is exact for degree $2N-1$ polynomials, we have that
\begin{align*}
\bm{v}_q^T\bm{Q}_i \bm{u}_q &= \bm{v}_q^T\bm{W} \bm{D}^i \bm{P}_q \bm{u}_q = \int_{\hat{D}} \pd{\Pi_N u}{\hat{x}_i} v = \int_{\partial \hat{D}} (\Pi_N u) v \hat{n}_i - \int_{\hat{D}} \LRp{\Pi_N u} \pd{v}{\hat{x}_i}.
\end{align*}
Since the volume integrand $\LRp{\Pi_N u} \pd{v}{\hat{x}_i} \in V^{2N-1}$, it is exactly integrated using the volume quadrature.  Additionally, since $\Pi_N u \in V^N$ and $v\in V^M$ on the surface of $\hat{D}$, the surface quadrature is exact for the surface integral and 
\begin{align*}
\int_{\partial \hat{D}} (\Pi_N u) v \hat{n}_i -   \int_{\hat{D}} \LRp{\Pi_N u} \pd{v}{\hat{x}_i} &= \bm{v}_f^T\bm{W}_f{\rm diag}(\hat{\bm{n}}_i) \bm{V}_f\bm{P}_q\bm{u}_q - \LRp{\bm{V}_q\bm{D}^i\bm{P}_q\bm{v}_q}^T\bm{W} \bm{V}_q\bm{P}_q\bm{u}_q\\
&= \bm{v}_q^T \LRp{\bm{V}_f\bm{P}_q}^T\bm{W}_f{\rm diag}(\hat{\bm{n}}_i) \bm{V}_f\bm{P}_q\bm{u}_q - \bm{v}_q^T \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T \bm{W} \bm{V}_q\bm{P}_q\bm{u}_q,
\end{align*}
where we have used that, since $v\in V^N$, $\bm{v}_f = \bm{V}_f\bm{P}_q\bm{v}_q$.  The proof is completed by noting that 
\[
\bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T \bm{W} \bm{V}_q\bm{P}_q = \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{M}\bm{P}_q = \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T\bm{W} = \bm{Q}_i^T.
\]
\end{proof}

>>>>

%\documentclass[preprint,10pt]{elsarticle}
%\documentclass[preprint,10pt]{article}
\documentclass[review]{siamart0216}
%\documentclass{siamart0216}

%\usepackage{fullpage}
%\usepackage[colorlinks=true]{hyperref}

\usepackage{amsmath,amssymb,amsfonts}
%\usepackage{amsthm}
%\theoremstyle{definition}
%\newtheorem{definition}{Definition}
%\theoremstyle{lemma}
%\newtheorem{lemma}{Lemma}
%\theoremstyle{corollary}
%\newtheorem{corollary}{Corollary}
%\theoremstyle{theorem}
%\newtheorem{theorem}{Theorem}
\newtheorem*{remark}{Remark}
\theoremstyle{assumption}
\newtheorem{assumption}{Assumption}

@@

\date{}
\author{Jesse Chan}
\title{Skew-symmetric entropy stable discontinuous Galerkin formulations with application to hybrid meshes }

\graphicspath{{./figs/}}

@@

\maketitle

%\begin{frontmatter}
%\title{Entropy stable skew-symmetric discontinuous Galerkin formulations with applications to hybrid and non-conforming meshes. }
%
%\author[rice]{Jesse Chan\corref{cor1}}
%\ead{Jesse.Chan@caam.rice.edu}
%\address[rice]{Department of Computational and Applied Mathematics, Rice University, 6100 Main St, Houston, TX, 77005}
%
%\begin{abstract}
%Entropy stable high order methods for nonlinear conservation laws satisfy an inherent discrete entropy inequality.  
%\end{abstract}
%\end{frontmatter}

@@

High order methods for the simulation of time-dependent compressible flow have the potential to achieve higher levels of accuracy at lower costs compared to current low order schemes \cite{wang2013high}.  In addition to superior accuracy and efficiency on modern computing architectures, the low numerical dispersion and dissipation of high order methods \cite{ainsworth2004dispersive} enables the accurate propagation of waves over long distances and time scales.  The same properties also make high order methods attractive for the resolution of unsteady phenomena such as vorticular and turbulent flows, which are sensitive to numerical dissipation \cite{visbal1999high, wang2013high}.  However, while high order methods are often provably stable for wave problems, high order schemes for nonlinear conservation laws have consistently been hampered by problems of instability.

When applied to nonlinear conservation laws, high order methods can experience artificial growth and blow-up near under-resolved features such as shocks, turbulence, or boundary layers.  In practice, the application of high order methods to practical problems requires shock capturing and stabilization techniques (such as artificial viscosity) or solution regularization (such as filtering or limiting) to prevent solution blow-up.  The resulting schemes for nonlinear conservation laws walk a fine line between stability, robustness, and accuracy.  Aggressive stabilization or regularization can result in the loss of high order accuracy, while too little can result in instability \cite{wang2013high}.  Moreover, it can be difficult to determine robust expressions for stabilization paramaters, as parameters which work for one simulation can fail when applied to a different physical regime or discretization setting.

These issues have motivated the introduction of high order \textit{entropy stable} discretizations, which satisfy a semi-discrete entropy inequality while maintaining high order accuracy in smooth regions.  Proofs of continuous entropy inequalities rely on the chain rule.  In contrast, these discrete entropy inequalities account for the loss of the chain rule due to effects such as quadrature errors, which are incurred when applying polynomially exact quadrature rules to nonlinear and rational integrands within the DG formulation.  These schemes were first introduced as high order collocation methods on tensor product elements in \cite{fisher2013high, carpenter2014entropy, gassner2016split, gassner2017br1}, and were extended to simplicial elements in \cite{crean2017high, chen2017entropy, crean2018entropy, chan2017discretely}.  These methods have also been extended to staggered grid methods \cite{parsani2016entropy} and non-conforming meshes \cite{friedrich2017entropy}.  Entropy stable boundary conditions have also been determined for the compressible Euler and Navier-Stokes equations \cite{parsani2015entropy, svard2018entropy}.

@@

The schemes in \cite{chan2017discretely} begin on the reference element $\hat{D}$ by approximating the solution in a degree $N$ polynomial basis $\LRc{\phi_j({\bm{x}})}_{i=1}^{N_p}$.  It is assumed that volume and surface quadrature rules $({\bm{x}}_i, w_i)$, $\LRp{{\bm{x}}^f_i,w^f_i}$ are provided for $\hat{D}$ and the surface $\partial \hat{D}$ of $\hat{D}$, respectively.  We make the following assumption on the volume quadrature rule:
\begin{assumption}
The volume quadrature rule integrates exactly all polynomials up to degree $(2N-1)$.
\label{ass:quad}
\end{assumption}
In addition to being related to the positive definiteness of the mass matrix, Assumption~\ref{ass:quad} ensures that, for any $u,v \in V^N$, the following integral is exactly approximated using quadrature
\[
\int_{\hat{D}} \pd{u(\hat{\bm{x}})}{\hat{x}_i} v(\hat{\bm{x}})\diff{\hat{\bm{x}}} = \sum_{k=1}^{N_q} w_k \pd{u(\hat{\bm{x}}_k)}{\hat{x}_i} v(\hat{\bm{x}}_k).  
\]
In other words, evaluating the integral of the derivative of $u$ and $v$ using quadrature is exact because the integrand is a polynomial of degree $(2N-1)$.

\note{TODO: Standardize hat notation for operators/normals: can refer to operators on reference element and do $k$ superscript for physical elements, or do hats on reference element.}

@@

Let $\bm{n}$ denote the outward unit normal $\bm{n}$ on $\hat{D}$, and let $\bm{n}_i$ denote the vector containing values of the $i$th component at all surface quadrature points.  The ``decoupled'' summation by parts operator $\bm{D}^i_N$ is defined as the block matrix involving both volume and surface quadratures
\begin{align}
\bm{D}^i_N  &= \LRs{
\begin{array}{cc}
\bm{D}^i_q - \frac{1}{2}\bm{V}_q \bm{L}_f {\rm diag}({\bm{n}}_i) \bm{V}_f\bm{P}_q &  \frac{1}{2}\bm{V}_q\bm{L}_f{\rm diag}({\bm{n}}_i)\\
-\frac{1}{2}{\rm diag}({\bm{n}}_i)\bm{V}_f\bm{P}_q & \frac{1}{2}{\rm diag}({\bm{n}}_i)
\end{array}}.  
\label{eq:DN}
\end{align}
When combined with projection and lifting matrices, $\bm{D}^i_N$ produces a high order polynomial approximation of $f\pd{g}{x_i}$. 
Let $f, g$ be differentiable functions, and let $\bm{f}_i = f(\bm{x}_i)$, $\bm{g}_i = g(\bm{x}_i)$ denote values of $f,g$ at volume and surface quadrature points.  Then,
\begin{gather}
f\pd{g}{x_i} \approx \LRs{\begin{array}{cc}
\bm{P}_q & \bm{L}_f\end{array}} {\rm diag}\LRp{\bm{f}}\bm{D}^i_N \bm{g}.  
\label{eq:dnapprox}
\end{gather}
The approximation can also be interpreted as solving a variational problem.  Let $n_i$ be the $i$th component of the unit normal on $\hat{D}$.  Then (\ref{eq:dnapprox}) is equivalent to finding $u(\bm{x}) \approx f\pd{g}{x_i}$ such that 
\[
\int_{\hat{D}} u(\bm{x})v(\bm{x})  = \int_{\hat{D}}{g\pd{\Pi_Nf}{x_i}v} + \int_{\partial \hat{D}}{(f-\Pi_Nf)\frac{\LRp{gv + \Pi_N(gv)}}{2}}n_i, \qquad \forall v\in V^N\LRp{\hat{D}}.
\]
If the surface quadrature rule is exact for polynomials of degree $2N$, then the decoupled SBP operator $\bm{D}^i_N$ also satisfies a summation by parts property.  
\begin{gather}
\bm{Q}^i_N = \bm{W}_N \bm{D}^i_N, \qquad \bm{W}_N = \LRp{\begin{array}{cc}\bm{W}&\\ & \bm{W}_f\end{array}},\label{eq:dsbp}\\ \bm{Q}^i_N+\LRp{\bm{Q}^i_N}^T = \LRp{\begin{array}{cc}\bm{0}&\\ & \bm{W}_f {\rm diag}(\bm{n}_i)\end{array}} = \bm{B}_N\nonumber
\end{gather}
By noting that $\bm{W}\bm{V}_q \bm{L}_q = \bm{P}_q^T\bm{V}_f^T\bm{W}_f$, the entries of $\bm{Q}^i_N$ can be re-written as
\begin{align}
\bm{Q}_i &= \bm{W}\bm{V}_q\bm{D}^i\bm{P}_q,\\
\bm{Q}^i_N  &= \LRs{
\begin{array}{cc}
\bm{Q}_i - \frac{1}{2}\LRp{\bm{V}_f \bm{P}_q}^T  \bm{W}_f{\rm diag}({\bm{n}}_i) \bm{V}_f\bm{P}_q &  \frac{1}{2}\LRp{\bm{W}_f{\rm diag}({\bm{n}}_i)\bm{V}_f\bm{P}_q}^T\\
-\frac{1}{2}\bm{W}_f{\rm diag}({\bm{n}}_i)\bm{V}_f\bm{P}_q & \frac{1}{2}\bm{W}_f{\rm diag}({\bm{n}}_i)
\end{array}}.
\label{eq:QN}
\end{align}
from which the SBP property (\ref{eq:dsbp}) is more apparent.

@@

\subsection{Loss of the SBP property for hybrid and non-conforming meshes}

While the SBP property has been used to derive entropy stable schemes, the SBP property is difficult or impossible to enforce in certain discretization settings, such as hybrid and non-conforming meshes arises.  This difficulty is a result of the choices of surface quadrature which naturally arise for these settings.

\begin{figure}
\centering
\begingroup
\captionsetup[subfigure]{width=.425\textwidth}
\subfloat[Insufficiently accurate surface quadrature on the triangle element.]{\includegraphics[width=.425\textwidth]{figs/hybrid2D.png}\label{subfig:hybrid1}}
\endgroup
\hspace{2em}
\subfloat[Incompatible surface quadrature on the quadrilateral element.]{\includegraphics[width=.425\textwidth]{figs/hybrid2D_GQ.png}\label{subfig:hybrid2}}
\caption{Examples of interface couplings which do not result in an SBP property and are not entropy stable.  On the left, the surface quadrature is a $(N+1)$ point GLL rule, and results in a loss of the SBP property on the triangle.  On the right, the surface quadrature is a $(N+1)$ point Gauss-Legendre rule, and results in a loss of the SBP property on the quadrilateral.  }
\label{fig:hybrid}
\end{figure}

@@

\begin{figure}[!h]
\centering
\begingroup
\captionsetup[subfigure]{width=.45\textwidth}
\subfloat[Entropy stable inter-element coupling in \cite{friedrich2017entropy} for a non-conforming interface.]{\raisebox{0em}{\includegraphics[width=.45\textwidth]{figs/nonconSBP.png}}\label{subfig:noncon2}}
\hspace{2em}
\subfloat[Entropy stable inter-element coupling in this work for a non-conforming interface.  The dotted black lines denote communication between neighboring elements.]{\raisebox{-0em}{\includegraphics[width=.45\textwidth]{figs/nonconQuad.png}}\label{subfig:noncon1}}
\endgroup
%\subfloat[3D hexahedra-pyramid coupling]{\includegraphics[width=.31\textwidth]{figs/hybrid3D.png}}
\caption{Illustration of two different entropy stable inter-element couplings for $h$ non-conforming meshes.  Each dashed red line indicates a dyadic flux computation required between two nodes. } %The coupling terms in \cite{friedrich2017entropy} requires computing dyadic fluxes between \emph{each} pair of nodes on adjacent interfaces.  The new approach results in a simpler communication pattern between elements. }
\label{fig:noncon}
\end{figure}

@@

\begin{lemma}
Let $v\in V^N$, and let the surface trace $\LRl{v}_{\partial \hat{D}} \in V^M\LRp{\partial \hat{D}}$ for some $M \leq N$. If the surface quadrature is exact for polynomials of degree $N+M$,  then for an arbitrary vector $\bm{u}_q$, 
\[
\bm{v}_q^T\bm{Q}_i \bm{u}_q = \bm{v}_q^T\LRp{ \LRp{\bm{V}_f\bm{P}_q}^T \bm{W}_f \diag{{\bm{n}}_i}\bm{V}_f\bm{P}_q - \bm{Q}_i^T}\bm{u}_q
\]
where $\bm{v}$ denotes the vector of values of $v$ of at both volume and surface quadrature points. 
\label{lemma:vsbp}
\end{lemma}
\begin{proof}
Recall that $\bm{Q}_i = \bm{W} \bm{V}_q \bm{D}^i\bm{P}_q$.  Since the volume quadrature is exact for degree $2N-1$ polynomials, we have that
\begin{align*}
\bm{v}_q^T\bm{Q}_i \bm{u}_q &= \bm{v}_q^T\bm{W} \bm{D}^i \bm{P}_q \bm{u}_q = \int_{\hat{D}} \pd{\Pi_N u}{\hat{x}_i} v = \int_{\partial \hat{D}} (\Pi_N u) v \hat{n}_i - \int_{\hat{D}} \LRp{\Pi_N u} \pd{v}{\hat{x}_i}.
\end{align*}
Since the volume integrand $\LRp{\Pi_N u} \pd{v}{\hat{x}_i} \in V^{2N-1}$, it is exactly integrated using the volume quadrature.  Additionally, since $\Pi_N u \in V^N$ and $v\in V^M$ on the surface of $\hat{D}$, the surface quadrature is exact for the surface integral and 
\begin{align*}
&\int_{\partial \hat{D}} (\Pi_N u) v \hat{n}_i -   \int_{\hat{D}} \LRp{\Pi_N u} \pd{v}{\hat{x}_i} =\\
 &\bm{v}_f^T\bm{W}_f{\rm diag}(\hat{\bm{n}}_i) \bm{V}_f\bm{P}_q\bm{u}_q - \LRp{\bm{V}_q\bm{D}^i\bm{P}_q\bm{v}_q}^T\bm{W} \bm{V}_q\bm{P}_q\bm{u}_q\\
=& \bm{v}_q^T \LRp{\bm{V}_f\bm{P}_q}^T\bm{W}_f{\rm diag}(\hat{\bm{n}}_i) \bm{V}_f\bm{P}_q\bm{u}_q - \bm{v}_q^T \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T \bm{W} \bm{V}_q\bm{P}_q\bm{u}_q,
\end{align*}
where we have used that, since $v\in V^N$, $\bm{v}_f = \bm{V}_f\bm{P}_q\bm{v}_q$.  The proof is completed by noting that 
\[
\bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T \bm{W} \bm{V}_q\bm{P}_q = \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{M}\bm{P}_q = \bm{P}_q^T\LRp{\bm{D}^i}^T \bm{V}_q^T\bm{W} = \bm{Q}_i^T.
\]
\end{proof}

The variational SBP property also extends to the decoupled SBP operator.  This property, along with the exact differentiation of constants, is necessary for the proof of entropy stability.  
\begin{lemma}
Let $\bm{D}^i_N$ be a decoupled SBP operator on the reference element $\hat{D}$, and let the surface quadrature be exact for polynomials of degree $N+M$ for some $M \leq N$.  Suppose $v\in V^N$ and that $\LRl{v}_{\partial \hat{D}} \in P^M\LRp{\partial \hat{D}}$, Then, 
\[
\bm{v}^T\bm{Q}^i_N\bm{u} = \bm{v}^T\LRp{\bm{B}^i_N - \LRp{\bm{Q}^i_N}^T}\bm{u}.%, \qquad \bm{Q}^i_N \bm{1} = \bm{0}.
\]
where $\bm{v}$ denotes the values of $v$ of at both volume and surface quadrature points.  
\label{lemma:vdsbp}
\end{lemma}
\begin{proof}
For convenience, let $\bm{u}_q, \bm{u}_f$ denote evaluations of $u$ at volume and surface points, such that 
\[
\bm{u} = \begin{pmatrix} \bm{u}_q\\ \bm{u}_f\end{pmatrix}, \qquad \bm{v} = \begin{pmatrix} \bm{v}_q\\ \bm{v}_f\end{pmatrix}.  
\]
The proof of the variational summation by parts property uses the definition of $\bm{Q}^i_N$ (\ref{eq:QN}), 
\begin{align*}
\bm{v}^T\bm{Q}^i_N\bm{u} &= \bm{v}_q^T\bm{Q}_i \bm{u}_q - \frac{1}{2}\LRp{\bm{V}_f\bm{P}_q\bm{v}_q}^T\bm{W}_f\hat{\bm{n}} \LRp{\bm{V}_f\bm{P}_q\bm{u}_q} + \frac{1}{2}\LRp{\bm{V}_f\bm{P}_q\bm{v}_q}^T\bm{W}_f\hat{\bm{n}} \bm{u}_f\\
& - \frac{1}{2}\bm{v}_f^T\bm{W}_f\hat{\bm{n}} \LRp{\bm{V}_f\bm{P}_q\bm{u}_q} + \frac{1}{2}\bm{v}_f^T\bm{W}_f\hat{\bm{n}} \bm{u}_f.
\end{align*}
Applying Lemma~\ref{lemma:vsbp} then yields
\begin{align*}
\bm{v}^T\bm{Q}^i_N\bm{u} =& -\bm{v}_q^T\bm{Q}^T_i \bm{u}_q + \frac{1}{2}\LRp{\bm{V}_f\bm{P}_q\bm{v}_q}^T\bm{W}_f\hat{\bm{n}} \LRp{\bm{V}_f\bm{P}_q\bm{u}_q} + \frac{1}{2}\LRp{\bm{V}_f\bm{P}_q\bm{v}_q}^T\bm{W}_f\hat{\bm{n}} \bm{u}_f\\
&\qquad - \frac{1}{2}\bm{v}_f^T\bm{W}_f\hat{\bm{n}} \LRp{\bm{V}_f\bm{P}_q\bm{u}_q} - \frac{1}{2}\bm{v}_f^T\bm{W}_f\hat{\bm{n}} \bm{u}_f + \bm{v}_f^T\bm{W}_f\hat{\bm{n}} \bm{u}_f\\
=& \begin{pmatrix} \bm{v}_q\\ \bm{v}_f\end{pmatrix}^T 
\left(\begin{pmatrix}
\bm{0}& \\
& \bm{W}_f\diag{\hat{\bm{n}}}
\end{pmatrix}\right.\\
&+
\left.\begin{pmatrix}
-\bm{Q}_i^T + \frac{1}{2}\LRp{\bm{V}_f\bm{P}_q}^T\bm{W}_f\hat{\bm{n}} \bm{V}_f\bm{P}_q & -\frac{1}{2} \LRp{\bm{W}_f\hat{\bm{n}} \bm{V}_f\bm{P}_q}^T\\
\frac{1}{2}\bm{W}_f\hat{\bm{n}} \bm{V}_f\bm{P}_q & -\frac{1}{2}\bm{W}_f\hat{\bm{n}}
\end{pmatrix}  \right)
\begin{pmatrix} \bm{u}_q\\ \bm{u}_f\end{pmatrix}\\
=& \bm{v}^T\LRp{\bm{B}_N - \LRp{\bm{Q}^i_N}^T}\bm{u}.
\end{align*}
\end{proof}
Using Lemma~\ref{lemma:vdsbp}, we will determine the accuracy of the surface quadrature necessary to ensure the proof of semi-discrete entropy stability is valid.  This minimal degree of surface quadrature accuracy will depend on the nature of the reference-to-physical mapping (e.g.\ affine vs curved elements).

\begin{corollary}
\label{lemma:sbpcor}
Let the surface quadrature be exact for polynomials of degree $N$.  Then, 
\[
\bm{1}^T\bm{Q}^i_N\bm{u} = \bm{1}^T\bm{B}_N\bm{u}, \qquad \bm{Q}^i_N\bm{1} = \bm{0},
\]
where $\bm{u}$ is a vector of values of some function at volume and surface quadrature points. %is an arbitrary vector of appropriate size.  
\end{corollary}
\begin{proof}
The proof that $\bm{Q}^i_N \bm{1} = \bm{0}$ follows from the property that polynomials are equal to their $L^2$ projection, and is identical to that of \cite{chan2017discretely,chan2018discretely}.     The proof of the first equality follows from $\bm{Q}^i_N \bm{1} = \bm{0}$ and Lemma~\ref{lemma:vdsbp} with $M=0$
\[
\bm{1}^T\LRp{\bm{Q}^i_N}\bm{u} = \bm{1}^T\LRp{\bm{B}_N - \LRp{\bm{Q}^i_N}^T}\bm{u} = \bm{1}^T{\bm{B}_N}\bm{u}.
\]
\end{proof}
\subsection{Entropy stability on affine meshes}

The high order methods in this work ensure that (\ref{eq:entropyineq}) is satisfied discretely by avoiding the use of the chain rule in the proof of entropy dissipation.  These ``entropy stable'' schemes rely on two main ingredients: an entropy stable numerical flux as defined by Tadmor \cite{tadmor1987numerical} and a concept referred to as ``flux differencing''.  % entropy stable schemes in one-dimension.  
Let $\bm{f}_S\LRp{\bm{u}_L,\bm{u}_R}$ be a numerical flux function which is a function of ``left'' and ``right'' states $\bm{u}_L,\bm{u}_R$.  
The numerical flux $\bm{f}_S$ is \textit{entropy conservative} if it satisfies the following three conditions:  
%The numerical flux $\bm{f}_S$ is an \textit{entropy stable} flux function if it satisfies the following three conditions:  
\begin{gather}
\bm{f}_S(\bm{u},\bm{u}) = \bm{f}(\bm{u}), \qquad \text{(consistency)}\\
\bm{f}_S(\bm{u}_L,\bm{u}_R) = \bm{f}_S(\bm{u}_R,\bm{u}_R), \qquad \text{(symmetry)}\nonumber\\
\LRp{\bm{v}_L-\bm{v}_R}^T\bm{f}_S(\bm{u}_L,\bm{u}_R) = \psi(\bm{u}_L) - \psi(\bm{u}_R), \qquad \text{(conservation)}\nonumber
%\LRp{\bm{v}_L-\bm{v}_R}^T\bm{f}_S(\bm{u}_L,\bm{u}_R) \leq \psi(\bm{u}_L) - \psi(\bm{u}_R), \qquad \text{(entropy dissipation)}\nonumber
\label{eq:esflux}
\end{gather}
%If instead of the third condition, $\bm{f}_S$ satisfies a \emph{stability} property 
%\[
%\LRp{\bm{v}_L-\bm{v}_R}^T\bm{f}_S(\bm{u}_L,\bm{u}_R) \leq \psi(\bm{u}_L) - \psi(\bm{u}_R)$, the flux $\bm{f}_S$ is referred to as \textit{entropy stable}.
